---
title: "Introduction to moderncor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to moderncor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `moderncor` package provides a single unified interface for computing a wide variety of classical and modern correlation coefficients. This guide introduces the core features of the package.

## Installation and Setup

Once installed, you can load the package as follows:

```r
library(moderncor)
```

## Basic Usage with Vectors

Let's generate some synthetic data with a non-linear parabolic relationship where $y = x^2 + \epsilon$:

```{r setup-data}
set.seed(123)
x <- runif(100, -1, 1)
y <- x^2 + rnorm(100, sd = 0.1)
```

Because the relationship is non-linear and symmetric, classical Pearson correlation will fail to capture the dependence:

```r
library(moderncor)
moderncor(x, y, method = "pearson")
```

With `moderncor`, you can compute distance correlation (`dcor`) or Chatterjee's Xi correlation (`xi`) using the same interface to capture the non-linear relationship:

```r
# Distance Correlation (captures non-linear dependencies)
moderncor(x, y, method = "dcor")

# Chatterjee's Xi (captures functional dependence)
moderncor(x, y, method = "xi")
```

## Matrix and Data Frame Input

If you pass a matrix or a `data.frame` to `moderncor()`, it will compute the pairwise correlation matrix of the columns:

```r
# Compute Spearman correlation matrix for iris dataset
res_mat <- moderncor(iris[, 1:4], method = "spearman")
res_mat
```

## Tidy Output using `as.data.frame`

You can convert the output of `moderncor()` to a tidy data frame using `as.data.frame()`. This is particularly useful for correlation matrices:

```r
# Convert correlation matrix to tidy data frame
df <- as.data.frame(res_mat)
head(df)
```

This returns a data frame containing the variables being compared (`var1` and `var2`), the correlation coefficient (`r`), and p-values (`p.value`) if they were calculated.

## Controlling P-value Computation

For large datasets, calculating p-values for modern methods (such as MIC, HSIC, or Mutual Information) can be slow because they rely on permutation tests. You can disable p-value calculations by setting `p_value = FALSE` for a significant speedup:

```r
# Compute only the estimate, without p-values
moderncor(x, y, method = "dcor", p_value = FALSE)
```

## Querying Available Methods

To see all supported correlation methods and their required packages:

```r
available_methods()
```

To get details on a specific method:

```r
method_info("dcor")
```
